Supplementary Information

### 2020-09-11 doc/manuscript/MEE_manuscript.Rmd
integrated Zofin feedback

**Authors (not yet complete, final order TBD):**

Kristina J. Anderson-Teixeira^1,2^*,
Valentine Herrmann^1^,
Christy Rollinson^3^,
Bianca Gonzalez^1^,
Erika B. Gonzalez-Akre^1^,
Neil Pederson^4^,
Camille Piponiot^1,2^,
Ross Alexander,
Craig D. Allen,
Raquel Alfaro-Sánchez,
Tala Awada,
Jennifer Baltzar,
Patrick Baker,
Sarayudh Bunyavejchewin,
Paolo Cherubini,
Justin Cooper,
Stuart Davies^2^,
Ryan Helcoski^1^,
Jakub Kašpar^cz^,
Kamil Kral,
James Lutz,
Ellis Q. Margolis,
Justin Maxwell,
Sean McMahon^2^,
Helene Muller-Landau^2^,
Sabrina Russo,
Pavel Šamonil^cz^,
Anastasia Sniderhan,
Alan J. Tepley^1,#^,
Ivana Vašíčková^cz^, 
Mart Vlam,
Peter Zuidema

@@

X#. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada

@@

**Table 1 | Sites included in this analysis**. Here and throughout, sites are ordered by latitude. Additional site information is provided in Table S1, and species and their sample sizes are detailed in Tables S2-S3. *Climate will be filled in based on CRU data.*
![ ](tables_figures/sites_table.png)
```{r eval=FALSE, echo=FALSE}
# refs in this table need to go in the nocite field below
```
---
nocite: | 
  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @tumajer_increasing_2017; @alfaro-sanchez_growth_2017; @helcoski_growing_2019; @aus_de_ar_tree_2018; @bumann_assessing_2019
---

@@

![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table S2) to potential evapotranspiration ($PET$). Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables. Black rectangle represents the period selected by *climwin* as the most influential window. Panels (**b-d**) give statistics for time windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diaganol indicate single-month tests (akin to panel **a**). Panels (**b**) and (**c**) give values of linear and quadratic terms for each time window, and (**d**) gives the $\Delta AIC$ for each. The time window with the minimum $\Delta AIC$ (0-3 months prior to August, or May-July; black circles), was identified as the optimal window. Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*. Finally, panel (**f**) shows GLS model output, where $PET$ was a candidate driver variable (along with $PPT$; $DBH$ not included in this model). Plotted are responses of species for which $PET$ was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant. Transparent ribbons indicate 95% confidence intervals.](tables_figures/quilt_comparison.png)

@@

![**Figure 3 | Species-level responses of $\Delta r$ to climwin-selected variables in precipitation and temperature variable groups.** For each species (color-coded as in Fig. 4), relationships are plotted if included in top model. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD. (THIS FIGURE WILL PROBABLY CHANGED INTO A 4COL X 5 ROW MATIX, DROPPING SPECIES KEY, WHICH CAN BE SEEN IN OTHER FIGURES)](tables_figures/pre_temp_groups.png)

@@

![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** Relationships for species are plottend when included in the top model. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are signficant, DASHED lines when only one term is signficant, and dotted lines when neither is signficant. Transparent ribbons indicate 95% confidence intervals.](tables_figures/DBH_responses.png)

@@

![**Figure 5 | Comparison of full top models for each growth metric ($\Delta r$, $BAI$, $\Delta AGB$) at sites where climatic controls are strong (Little Tesuque, left panel) and modest (SCBI, right panel).** Plotted are best fit models for each species, with transparent ribbons indicating 95% confidence intervals. For climate variables, best-fit polynomials are plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/show_case_response_plots.png)

@@

@@

All this points to the potential importance of considering DBH in climate reconstructions.
Traditional dendro methods seek to remove the influence of $DBH$, but our finding of prevalent $DBH$-$\Delta r$ interactions indicate that it should not be ignored.
DBH is not always collected when cores are taken, and is not routinely preserved alongside tree-ring data. For example, the International Tree-Ring Data Bank (ITRDB) contains no structure for storing DBH records. DEN is new alternative [@rayback_dendroecological_2020].

>>>>

**Authors:**

[Kristina J. Anderson-Teixeira^1,2^*](0000-0001-8461-9713),
Valentine Herrmann^1^,
Christy Rollinson^3^,
Bianca Gonzalez^1^,
Erika B. Gonzalez-Akre^1^,
Neil Pederson^4^,
Camille Piponiot^1,2^,
Ross Alexander,
Craig D. Allen,
Raquel Alfaro-Sánchez,
Tala Awada,
Jennifer Baltzar,
Patrick Baker,
Sarayudh Bunyavejchewin,
Paolo Cherubini,
Justin Cooper,
Stuart Davies^2^,
Ryan Helcoski^1^,
[Jakub Kašpar^cz^](0000-0003-1780-6310),
Kamil Kral,
James Lutz,
Ellis Q. Margolis,
Justin Maxwell,
Sean McMahon^2^,
Helene Muller-Landau^2^,
Sabrina Russo,
[Pavel Šamonil^cz^](0000-0002-7722-8797),
Anastasia Sniderhan,
Alan J. Tepley^1,at^,
[Ivana Vašíčková^cz^](0000-0002-6070-5956), 
Mart Vlam,
Peter Zuidema

@@

at. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada

@@

**Table 1 | Sites included in this analysis**. Here and throughout, sites are ordered by latitude. Additional site information is provided in Table S1, and species and their sample sizes are detailed in Tables S2-S3. *Climate will be filled in based on CRU data.*
![ ](tables_figures/sites_table.png)
```{r eval=FALSE, echo=FALSE}
# refs in this table need to go in the nocite field below
```
---
nocite: | 
  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017; @helcoski_growing_2019; @aus_de_ar_tree_2018; @bumann_assessing_2019; @samonil_individual-based_2013; @kaspar_species-specific_nodate
---

@@

![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table S2) to potential evapotranspiration ($PET$). Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables. Black rectangle represents the period selected by *climwin* as the most influential window. Panels (**b-d**) give statistics for time windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diaganol indicate single-month tests (akin to panel **a**). Panels (**b**) and (**c**) give values of linear and quadratic terms for each time window, and (**d**) gives the $\Delta AIC$ for each. The time window with the minimum $\Delta AIC$ (0-3 months prior to August, or May-July; black circles), was identified as the optimal window. Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*. Finally, panel (**f**) shows GLS model output, where $PET$ was a candidate driver variable (along with $PPT$; $DBH$ not included in this model). Plotted are responses of species for which $PET$ was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals.](tables_figures/quilt_comparison.png)

@@

![**Figure 3 | Species-level responses of $\Delta r$ to climwin-selected variables in precipitation and temperature variable groups.** For each species (color-coded as in Fig. 4), relationships are plotted if included in top model. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD. (THIS FIGURE WILL PROBABLY CHANGED INTO A 4COL X 5 ROW MATIX, DROPPING SPECIES KEY, WHICH CAN BE SEEN IN OTHER FIGURES)](tables_figures/pre_temp_groups.png)

@@

![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** Relationships for species are plotted when included in the top model. Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant, DASHED lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals.](tables_figures/DBH_responses.png)

@@

![**Figure 5 | Comparison of full top models for each growth metric ($\Delta r$, $BAI$, $\Delta AGB$) at sites where climatic controls are strong (Little Tesuque, left panel) and modest (SCBI, right panel).** Plotted are best fit models for each species, with transparent ribbons indicating 95% confidence intervals. For climate variables, best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/show_case_response_plots.png)

@@

* species climate sensitivity models could be improved by fitting climwin individually be species.

@@

All this points to the potential importance of considering DBH in climate reconstructions.
Traditional dendro methods seek to remove the influence of $DBH$, but our finding of prevalent $DBH$-$\Delta r$ interactions indicate that it should not be ignored.
*(Indeed, this is probably one factor contributing to the non-stationarity of climate signals in tree-ring records [@wilmking_global_2020].)*
*(	Interactive effects of DBH and climate may suggest significant impact of different sampling strategy on results per individual site. -- e.g., stronger drought signal if just large trees are sampled. This is consistent with the dendro tradition of sampling canopy trees for strongest drought signals; however, our results suggest that temperature signals may be stronger in young trees.  )*
*(height vs exposure/ social status of trees)*

*Recommendations*

Need to record/ preserve DBH.  DBH is not always collected when cores are taken, and is not routinely preserved alongside tree-ring data. For example, the International Tree-Ring Data Bank (ITRDB) contains no structure for storing DBH records. DEN is new alternative [@rayback_dendroecological_2020].

Also need to record canopy positions.

Sampling small trees should be important for disentangling whether non-stationarity of climate signals is a size artifact or driven by something else, may also be better for resolving temperature signals.

### 2020-09-11 doc/manuscript/MEE_manuscript_SI.Rmd
integrated Zofin feedback

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
ss_table <- read.csv("tables_figures/qualitative_results_comparison.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(ss_table, format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
  add_header_above(c(" "," ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
  column_spec(3, width = "2.5cm") %>%
  column_spec(4, width = "2.5cm") %>%
  column_spec(5, width = "2.5cm") %>%
  column_spec(6, width = "2.5cm") %>%
  column_spec(7, width = "2cm") %>%
  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 9)

# refs in this table need to go in the nocite field below
```
---
nocite: | 
  @alfaro-sanchez_growth_2017; @vlam_temperature_2014; @helcoski_growing_2019; @maxwell_declining_2016; @aus_de_ar_tree_2018; @bumann_assessing_2019; @tumajer_increasing_2017; @sniderhan_growth_2016
---
Results from this study are the climate-only model. 
Where previous studies examined numerous climate variables or time windows [e.g., @helcoski_growing_2019], we focus on those most relevant to our findings. 
Beyond the methodological differences, original studies vary from this one and from one another in factors including
exact set of cores analyzed,
climate data sources,
time frame of analysis,
approaches to identifying candidate climate variables and windows (including whether this is done on a site or species level), 
methods for detrending and standardizing to build chronologies,
and whether the effects of temperature and precipitation are considered separately (original studies) or additively (this study).
An analysis standardizing all of these factors for four species is present in Appendix **S2** and Fig. **S1**.

>>>>

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
breakrow1<-10
breakrow2<-24
breakrow3<-32
col_width<-"2.5cm"
table_pt1 <- read.csv("tables_figures/qualitative_results_comparison.csv", 
                     colClasses=c("NULL", NA, NA, NA, NA, NA, NA), 
                     col.names=c("NULL", "species", "previously observed", "observed here", "previously observed", "observed here", "reference"), 
                     skip=0, nrows=breakrow1,
                     stringsAsFactors = FALSE, check.names = FALSE)
kable(table_pt1, format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
  add_header_above(c(" ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
  column_spec(2:5, width = col_width) %>%
  column_spec(6, width = "2cm") %>%
  pack_rows("Barro Colorado Island, Panama", 1, 3, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Huai Kha Khaeng, Thailand", 4, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
# refs in this table need to go in the nocite field below
```

\newpage
Table S5, cont.
```{r eval = TRUE, echo=FALSE, warning=FALSE}
table_pt2 <- read.csv("tables_figures/qualitative_results_comparison.csv", 
                     colClasses=c("NULL", NA, NA, NA, NA, NA, NA), 
                     col.names=c("NULL", "species", "previously observed", "observed here", "previously observed", "observed here", "reference"), 
                     skip=breakrow1, nrows=breakrow2-breakrow1,
                     stringsAsFactors = FALSE, check.names = FALSE)
kable(table_pt2, format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
  add_header_above(c(" ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
  column_spec(2:5, width = col_width) %>%
  column_spec(6, width = "2cm") %>%
  pack_rows("Smithsonian Conservation Biology Institute, Virginia, USA", 1, 14, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
# refs in this table need to go in the nocite field below
```

\newpage
Table S5, cont.
```{r eval = TRUE, echo=FALSE, warning=FALSE}
table_pt3 <- read.csv("tables_figures/qualitative_results_comparison.csv", 
                     colClasses=c("NULL", NA, NA, NA, NA, NA, NA), 
                     col.names=c("NULL", "species", "previously observed", "observed here", "previously observed", "observed here", "reference"), 
                     skip=breakrow2, nrows=breakrow3-breakrow2,
                     stringsAsFactors = FALSE, check.names = FALSE)
kable(table_pt3, format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
  add_header_above(c(" ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
  column_spec(2:5, width = col_width) %>%
  column_spec(6, width = "2cm") %>%
  pack_rows("Žofín Forest Dynamics Plot, Czech Republic", 4, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Scotty Creek, NW Territories, Canada", 8, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
# refs in this table need to go in the nocite field below
```
---
nocite: | 
  @alfaro-sanchez_growth_2017; @vlam_temperature_2014; @helcoski_growing_2019; @maxwell_declining_2016; @aus_de_ar_tree_2018; @bumann_assessing_2019; @tumajer_increasing_2017; @sniderhan_growth_2016; @kaspar_species-specific_nodate
---
Results from this study are the climate-only model. 
Where previous studies examined numerous climate variables or time windows [e.g., @helcoski_growing_2019], we focus on those most relevant to our findings. 
Beyond the methodological differences, original studies vary from this one and from one another in factors including
exact set of cores analyzed,
climate data sources,
time frame of analysis,
approaches to identifying candidate climate variables and windows (including whether this is done on a site or species level), 
methods for detrending and standardizing to build chronologies,
and whether the effects of temperature and precipitation are considered separately (original studies) or additively (this study).
An analysis standardizing all of these factors for four species is present in Appendix **S2** and Fig. **S1**.
